### 2016-01-15 03-design.tex
Some typos, one or another citation

Modern network technology allows access to the memory of remote machines
through the network cards (NICs) without involvement from the remote host
altogether~\cite{Hamada_infinibandtrade,Beck:2011:PER:2043535.2043537}.
This has opened up the development of
extremely efficient remote memory services, often in the form of DKV stores.
This project stores $\pi$ in a DKV store in the collective memory of the
cluster, where $\pi[i]+\pi^{sum}[i]$ is the value for key $i$. We decided not
to use any of the existing implementations of RDMA DKV stores, and in stead
build our own DKV store on top of the ib-verbs API~\cite{ib-verbs}. The
main consideration is that our use case is unusually simple in a number
of important aspects. The KV layout is static in the sense that there are
no inserts or deletes after the initial population, which allows a static
partitioning of KV pairs over the machines. The values are all of the same size,
a vector of $K+1$ floats. The access pattern is very regular, because the
computation is partitioned into stages that are separated by barriers; a
stage either reads values or it updates values that it 'owns', so there
is no read/write or write/write contention. The existing DKV stores incur
overhead for insert/delete flexibility, concurrency control, or variable-sized
values. Our use case allows us to do any operation in exactly one RDMA read
or RDMA write. The Herd paper~\cite{Kalia:2014:URE:2740070.2626299,} argues
that replacing RDMA reads by
an RPC to the server, followed by an unreliable write, is faster. However,
they show this only holds for small payload packets, up to 256B. A $\pi$
packet in our application is typically thousands to hundreds of thousands
of 4-byte floats.

>>>>

Modern network technology allows access to the memory of remote machines
through the network cards (NICs) without involvement from the remote host
altogether~\cite{Hamada_infinibandtrade,Beck:2011:PER:2043535.2043537}.
This has opened up the development of
extremely efficient remote memory services, often in the form of DKV stores.
This project stores $\pi$ in a DKV store in the collective memory of the
cluster, where $\pi[i]+\pi^{sum}[i]$ is the value for key $i$. We decided not
to use any of the existing implementations of RDMA DKV stores, and in stead
build our own DKV store on top of the ib-verbs API~\cite{ib-verbs}. The
main consideration is that our use case is unusually simple in a number
of important aspects. The KV layout is static in the sense that there are
no inserts or deletes after the initial population, which allows a static
partitioning of KV pairs over the machines. The values are all of the same size,
a vector of $K+1$ floats. The access pattern is very regular, because the
computation is partitioned into stages that are separated by barriers; a
stage either reads values or it updates values that it 'owns', so there
is no read/write or write/write contention. The existing DKV stores incur
overhead for insert/delete flexibility, concurrency control, or variable-sized
values. Our use case allows us to do any operation in exactly one RDMA read
or RDMA write. The Herd paper~\cite{Kalia:2014:URE:2740070.2626299} argues
that replacing RDMA reads by
an RPC to the server, followed by an unreliable write, is faster. However,
they show this only holds for small payload packets, up to 256B. A $\pi$
packet in our application is typically thousands to hundreds of thousands
of 4-byte floats.

### 2016-01-15 04-evaluation.tex
Some typos, one or another citation

Empirical results were obtained by performing experiments on the VU and TU/D
DAS5 clusters which consist of 68 and 48 compute nodes respectively. Each
compute node is equipped with a dual 8-core Intel Xeon E5-2630v3 CPU clocked
at 2.40GHz, 64GB of memory and 8TB of storage. Moreover, the compute nodes of
each site are interconnected by
FDR~InfiniBand. All of the experiments reported in this section utilized
publicly available graphs from the Stanford Large Network Dataset
Collection~(SNAP) \cite{snapnets}. Table~\ref{tab-datasets} lists the
collection of data sets used.

@@

In order to evaluate the horizontal scalability of the distributed
implementation we tested the system's performance across different cluster
sizes while holding the problem size constant. For this study, we used the
com-Friendster graph as it contains the largest number of vertices and edges.
%
Figure~\ref{fig-strong-scaling}-a presents the execution time of 2048 algorithm
iterations across multiple cluster sizes. The x-axis starts from 8 worker nodes
as the data set is too large to fit into the collective memory of a smaller
cluster.
%
As shown in the figure, the execution
time steadily decreases by increasing the cluster size.
A deeper analysis of the individual execution phases of the algorithm provides
insights into the scalability if its building blocks.
%
In addition to the total execution time, Figure~\ref{fig-strong-scaling}-a
presents the cumulative time spent in individual computational phases across
iterations. Moreover, Figure~\ref{fig-strong-scaling}-b presents the speedup
achieved for the same experiments reported in
Figure~\ref{fig-strong-scaling}-a. As is clearly shown, the dominant phase of
the execution is update\_pi.
The reported total time for each cluster size is significantly less than the
sum of the execution times of the individual phases. This is due to the
overlapping execution of the two most expensive phases, namely, update\_pi and
deploy\_mini-batch. Both of these phases initially gain significant speedup with the
addition of compute nodes. However, the speedup curve gradually slows down for
larger cluster sizes as the work granularity of each worker node decreases,
limiting their resource utilization.
%
The time spent in update\_beta remains relatively constant across cluster sizes
as it performs an insignificant amount of work compared to the synchronization
overhead of a collective MPI reduction operation contained within it.

@@

A study of a system's weak scaling aids in the assessment of the communication
and synchronization overheads associated with the management of large clusters.
Similarly, it can expose complex issues that hinder performance at scale such
as load imbalance. To fairly evaluate the algorithm's weak scaling behavior we
conducted several executions varying the cluster size and the number of latent
communities proportionally. This methodology ensures that each compute node
performs a relatively constant amount of computational work across all
configurations. However, the number and size of messages exchanged between the
nodes would vary significantly. Figure~\ref{fig-weak-scaling}-a presents the
average execution
time per algorithm iteration across different cluster sizes. On the other hand,
Figure~\ref{fig-weak-scaling}-b reports the set number of communities for each
cluster configuration. The relative change in the average execution time per
iteration is insignificant even though the communication intensity increases
for larger cluster sizes. This observation verifies that the system's overall
overhead is minimal. Additionally, the experimental results shows that the
implementation is capable of achieving good speedups provided the input problem
is large enough for the given cluster size.

@@

As discussed in Section~\ref{design-section}, the collective memory of all
worker nodes serves as the storage for the state of the computation. As such,
the rows of $\pi$ are equally and statically partitioned across all workers.
Given that $\pi$ accesses are random, a node in a cluster of size $C$ must
fetch $(C-1)/C$ of all read requests over the network. Therefore, large cluster
configurations exhibit higher bandwidth demands and are more sensitive
to network latency. To reduce the negative effects of network latency on the
computation, a pipelining scheme was devised to prefetch data dependencies.
Figure~\ref{fig-pipeline} presents the execution time of 1024 algorithm
iterations on a 64-node cluster with double-bufferring enabled and disabled.
Naturally, increasing the number of communities causes a proportional increase
in execution time. However, when double-buffering is enabled, some of the
incurred network latency is hidden by overlapping it with computation.
Moreover, since both computation time and network latency increase with larger
$K$, the benefit of pipelining increases. This can be observed through the
widening gap between both lines depicted in Figure~\ref{fig-pipeline}.

@@

%################################################
\subsection{Horizontal vs. Vertical Scalability}
\begin{figure*}[t] % [htb]
  \centering
  \epsfig{file=plots/hpc-cloud.eps, width=\textwidth}
  \caption{Performance comparison between the distributed implementation
  running on DAS5 and the multi-threaded solution on machine with 40-cores and
  1TB of RAM.}
  \label{fig-scale-up}
\end{figure*}
One of the main drawbacks of designing a distributed solution for a given
algorithm is the inherent complexity of communication and synchronization. A
slightly simpler approach would be developing a multi-threaded version and
running it on a machine with abundant memory and CPU cores. In such a context,
access to all of the algorithm's state would be an order of magnitude faster
than RDMA. Additionally, the overhead associated with synchronizing threads is
negligible compared to using MPI primitives. To evaluate the efficacy of both
approaches we utilized SURFsara's HPC Cloud system to instantiate a virtual
machine with 40 Intel Xeon E7-4850 cores clocked at 2.00GHz cores and 1TB of
memory. The physical machine underlying
the HPC Cloud system contains 40-CPU cores and does not oversubscribe
resources. Therefore, by provisioning all 40~cores we ensured that there is no
resource contention from other users of the system. Figure~\ref{fig-scale-up}
reports the execution time per algorithm iteration for two experimental setups.
%
First, Figure~\ref{fig-scale-up}-a shows the performance of executing the
algorithm on the HPC Cloud
system with 40 and 16 cores compared to a single DAS5 node with 16-cores. This
test uses the com-DBLP data set to enable the use of a large number of
communities without running out of memory. It is clear from the results that
the performance can benefit from the additional cores provided by HPC Cloud system.
%
Further, Figure~\ref{fig-scale-up} tests the performance of the HPC Cloud
system compared to 64-nodes of the DAS5
cluster using the com-Friendster data set. Clearly, the parallel and
distributed implementation vastly
outperforms the single-node multi-threaded solution. Moreover, the trajectory
of both curves shows a widening gap between them suggesting that the relative
performance difference will increase for larger $K$.
%
In conclusion, the overhead of network communication in the distributed version
is more than compensated by the increasing compute power compared to
a single-node implementation.

@@

The previous sections focused on the computational performance of the
distributed implementation. However, the algorithm's throughput does not
necessarily indicate how fast it can converge to a solution. More specifically, it
remains unclear how many iterations are needed for the algorithm to reach a
stable state and terminate. Therefore, we now shift our focus to the
convergence time to assess the system's utility. Figure~\ref{fig-ppx} presents
the convergence time of the data sets listed in Table~\ref{tab-datasets}.
%
The results in sub-figures a, b and c where obtained by using 64 compute nodes
of DAS5. In Figure~\ref{fig-ppx}-a, the number of communities was set to 12K
which fully occupied the aggregate memory resources of all 64 machines since
com-Friendster has roughly 65 million vertices. In this case, the algorithm
reached a stable state after 3-4 hours. Next, Figures~\ref{fig-ppx}-b and
\ref{fig-ppx}-c present the convergence of com-LiveJournal and com-Orkut
respectively. As the number of vertices in these data sets is an order of
magnitude smaller than com-Friendster, we used a larger number of
communities. Naturally, the convergence time was extended as the complexity of
the algorithm increases dramatically with larger K. However, the system was
still capable of reporting results after \textbf{XXX time}.
%
Figures~\ref{fig-ppx}-d, \ref{fig-ppx}-e and \ref{fig-ppx}-f were all
configured to use the total number of ground-truth communities associated with
their respective data sets, com-Youtube, com-DBLP and com-Amazon. However,
these experiments were conducted on 14, 24 and 24 cluster nodes respectively.

>>>>

Empirical results were obtained by performing experiments on the VU and
Leiden University
DAS5 clusters which consist of 68 and 24 compute nodes respectively. Each
compute node is equipped with a dual 8-core Intel Xeon E5-2630v3 CPU clocked
at 2.40GHz, 64GB of memory and 8TB of storage. Moreover, the compute nodes of
each site are interconnected by
FDR~InfiniBand. The MPI implementation used is MVAPICH2, which is optimally
tuned for Infiniband~\cite{mvapich}. All of the experiments reported in this section utilized
publicly available graphs from the Stanford Large Network Dataset
Collection~(SNAP) \cite{snapnets}. Table~\ref{tab-datasets} lists the
collection of data sets used.

@@

In order to evaluate the horizontal scalability of the distributed
implementation we tested the system's performance across different cluster
sizes while holding the problem size constant. For this study, we used the
com-Friendster graph as it contains the largest number of vertices and edges.
%
Figure~\ref{fig-strong-scaling}-a presents the execution time of 2048 algorithm
iterations across multiple cluster sizes. The x-axis starts from 8 worker nodes
as the data set is too large to fit into the collective memory of a smaller
cluster.
%
As shown in the figure, the execution
time steadily decreases by increasing the cluster size.
A deeper analysis of the individual execution phases of the algorithm provides
insights into the scalability of its building blocks.
%
In addition to the total execution time, Figure~\ref{fig-strong-scaling}-a
presents the cumulative time spent in individual computational phases across
iterations. Moreover, Figure~\ref{fig-strong-scaling}-b presents the speedup
achieved for the same experiments reported in
Figure~\ref{fig-strong-scaling}-a. As is clearly shown, the dominant phase of
the execution is \textit{update\_pi}.
The reported total time for each cluster size is significantly less than the
sum of the execution times of the individual phases. This is due to the
overlapping execution of the two most expensive phases, namely,
\textit{update\_pi} and
\textit{deploy\_minibatch}. Both of these phases initially gain significant speedup with the
addition of compute nodes. However, the speedup curve gradually slows down for
larger cluster sizes as the work granularity of each worker node decreases,
limiting their resource utilization.
%
The time spent in \textit{update\_beta} remains relatively constant across cluster sizes
as it performs an insignificant amount of work compared to the synchronization
overhead of a collective MPI reduction operation contained within it.

@@

A study of a system's weak scaling aids in the assessment of the communication
and synchronization overheads associated with the management of large clusters.
Similarly, it can expose complex issues that hinder performance at scale such
as load imbalance. To fairly evaluate the algorithm's weak scaling behavior we
conducted several executions varying the cluster size and the number of latent
communities proportionally. This methodology ensures that each compute node
performs a relatively constant amount of computational work across all
configurations. However, the number and size of messages exchanged between the
nodes would vary significantly. Figure~\ref{fig-weak-scaling}-a presents the
average execution
time per algorithm iteration across different cluster sizes. On the other hand,
Figure~\ref{fig-weak-scaling}-b reports the number of communities for each
cluster configuration. The relative change in the average execution time per
iteration is insignificant even though the communication intensity increases
for larger cluster sizes. This observation verifies that the system's overall
overhead is minimal. Additionally, the experimental results shows that the
implementation is capable of achieving good speedups provided the input problem
is large enough for the given cluster size.

@@

As discussed in Section~\ref{design-section}, the collective memory of all
worker nodes serves as the storage for the state of the computation. As such,
the rows of $\pi$ are equally and statically partitioned across all workers.
Given that $\pi$ accesses are random, a node in a cluster of size $C$ must
fetch $(C-1)/C$ of all read requests over the network. Therefore, large cluster
configurations exhibit higher bandwidth demands and are more sensitive
to network latency. To reduce the negative effects of network latency on the
computation, a pipelining scheme was devised to prefetch data dependencies.
Figure~\ref{fig-pipeline} presents the execution time of 1024 algorithm
iterations on a 64-node cluster with double-buffering enabled and disabled.
Naturally, increasing the number of communities causes a proportional increase
in execution time. However, when double-buffering is enabled, some of the
incurred network latency is hidden by overlapping it with computation.
Moreover, since both computation time and network latency increase with larger
$K$, the benefit of pipelining increases. This can be observed through the
widening gap between both lines depicted in Figure~\ref{fig-pipeline}.

@@

%################################################
\subsection{Horizontal vs. Vertical Scalability}
\begin{figure*}[t] % [htb]
  \centering
  \epsfig{file=plots/hpc-cloud.eps, width=\textwidth}
  \caption{Performance comparison between the distributed implementation
  running on DAS5 and the multi-threaded solution on machine with 40-cores and
  1TB of RAM.}
  \label{fig-scale-up}
\end{figure*}
One of the main drawbacks of designing a distributed solution for a given
algorithm is the inherent complexity of communication and synchronization. A
considerably simpler approach would be developing a multi-threaded version and
running it on a machine with abundant memory and CPU cores. In such a context,
access to all of the algorithm's state would be an order of magnitude faster
than RDMA. Additionally, the overhead associated with synchronizing threads is
negligible compared to using MPI primitives. To evaluate the efficacy of both
approaches we utilized SURFsara's HPC Cloud system to instantiate a virtual
machine with 40 Intel Xeon E7-4850 cores clocked at 2.00GHz cores and 1TB of
memory. The physical machine underlying
the HPC Cloud system contains 40~CPU cores and does not oversubscribe
resources. Therefore, by provisioning all 40~cores we ensured that there is no
resource contention from other users of the system. Figure~\ref{fig-scale-up}
reports the execution time per algorithm iteration for two experimental setups.
%
First, Figure~\ref{fig-scale-up}-a shows the performance of executing the
algorithm on the HPC Cloud
system with 40 and 16 cores compared to a single DAS5 node with 16~cores. This
test uses the com-DBLP data set to enable the use of a large number of
communities without running out of memory. It is clear from the results that
the performance can benefit from the additional cores provided by HPC Cloud system.
%
Further, Figure~\ref{fig-scale-up} tests the performance of the HPC Cloud
system compared to 64~nodes of the DAS5
cluster using the com-Friendster data set. Clearly, the parallel and
distributed implementation vastly
outperforms the single-node multi-threaded solution. Moreover, the trajectory
of both curves shows a widening gap between them suggesting that the relative
performance difference will increase for larger $K$.
%
In conclusion, the overhead of network communication in the distributed version
is more than compensated by the increasing compute power compared to
a single-node implementation.

@@

The previous sections focused on the computational performance of the
distributed implementation. However, the algorithm's throughput does not
necessarily indicate how fast it can converge to a solution. More specifically, it
remains unclear how many iterations are needed for the algorithm to reach a
stable state and terminate. Therefore, we now shift our focus to the
convergence time to assess the system's utility. Figure~\ref{fig-ppx} presents
the convergence time of the data sets listed in Table~\ref{tab-datasets}.
%
The results in sub-figures a, b and c were obtained by using 64 compute nodes
of DAS5. In Figure~\ref{fig-ppx}-a, the number of communities was set to 12K
which fully occupied the aggregate memory resources of all 64 machines since
com-Friendster has roughly 65 million vertices. In this case, the algorithm
reached a stable state after 3-4 hours. Next, Figures~\ref{fig-ppx}-b and
\ref{fig-ppx}-c present the convergence of com-LiveJournal and com-Orkut
respectively. As the number of vertices in these data sets is an order of
magnitude smaller than com-Friendster, we used a larger number of
communities. Naturally, the convergence time was extended as the complexity of
the algorithm increases dramatically with larger K. However, the system was
capable of reporting results for both after 40 hours.
%
Figures~\ref{fig-ppx}-d, \ref{fig-ppx}-e and \ref{fig-ppx}-f were all
configured to use the total number of ground-truth communities associated with
their respective data sets, com-Youtube, com-DBLP and com-Amazon. However,
these experiments were conducted on 14, 24 and 24 cluster nodes respectively.
